### 2022-12-24 draft/draft.Rmd
fixed a bunch of typos

Graph and network data have become increasingly widespread in various fields including sociology, neuroscience, biostatistics, and computer science. 
This has resulted in challenges for researchers who rely on traditional statistical and machine learning methods that are incompatible with graph data and instead assume that the data exist as feature vectors. 
To illustrate this, consider the typical approach to building a statistical or machine learning model. 
Data are often represented as an $n \times p$ matrix $X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top$ in which each row $x_i \in \mathbb{R}^p$ is an observation of $p$ features and each column is a set of $n$ feature measurements. 
An analysis task for these data might be to come up with a classification model $\phi : \mathbb{R}^p \to \{1, 2, ..., K\}$ that uses the numerical values of each feature of a vector $x_i$ to calculate a predicted label $z_i \in \{1, 2, ..., K\}$. 
For instance, $\phi(x)$ might first compute the distances from $x$ to $K$ points in $\mathbb{R}^p$ and then assign $x$ to the label of the nearest point. 
Examples of this include linear discriminant analysis (in the case of supervised learning) and Lloyd's algorithm \citep{1056489} or Gaussian mixture models \citep{doi:10.1198/016214502760047131} (in the case of unsupervised learning). 
However, it is not obvious how this method would translate to data that are represented as graphs, in which observations consist of relationships among a set of objects rather than numerical attributes associated with each object: 
Instead of feature vector $x_i = \Big[ x_{i1} \mid \cdots \mid x_{ip} \Big]^\top \in \mathbb{R}^p$, we observe $a_i = \Big[ a_{i1} \mid \cdots \mid a_{in} \Big]^\top \in \mathbb{R}^n$ in which each $a_{ij}$ is object $i$'s relationship to object $j$. 
For such data, it is often necessary to alter existing algorithms, transform the graph data into Euclidean data, often called *graph embedding* or *spectral clustering* \citep{vonLuxburg2007}, and apply algorithms for Euclidean data on the embedding, or come up with new algorithms altogether.

\begin{example}[Graph clustering]
$K$-means clustering \citep{MacQueen1967} is a nonparametric clustering method that minimizes the objective function
\begin{equation}
\label{eq:kmeans}
\argmin_{z_1, ..., z_n} ~ W(z_1, ..., z_n; x_1, ..., x_n) = 
\argmin_{z_1, ..., z_n} ~ \sum_{k = 1}^K \sum_{x_i : z_i = k} \|x_i - \bar{x}_k\|_2^2,
\end{equation}
in which $x_1, ..., x_n \in \mathbb{R}^p$ are feature vectors of a sample, $z_1, ..., z_n \in \{1, ..., K\}$ are cluster assignments which we wish to optimize, and $\bar{x}_1, ..., \bar{x}_K \in \mathbb{R}^p$ are the centroids of each cluster, which act as nuisance parameters. 
A popular implementation of $K$-means clustering is Lloyd's algorithm \citep{1056489}, which is a type of coordinate descent algorithm in which at each iteration the cluster labels are updated according to the label of its nearest centroid and centroids are updated as the sample means of the points in each cluster.

If we have a graph instead of vectors in $\mathbb{R}^p$, it is not obvious how to translate $K$-means clustering to these data. 
In particular, there is no inherent notion of centroid or sample mean for graph data. 
However, there are inherent notions of distances among vertices, such as shortest path distance, expected commute time, or resistance distance. 
One way to adapt $K$-means to graph data is to use an equivalent objective function as equation \eqref{eq:kmeans} that doesn't use centroids (which are just nuisance parameters) and is stated only in terms of distances:  
$$
\argmin_{z_1, ..., z_n} ~ \tilde{W}(z_1, ..., z_n; x_1, ..., x_n) =
\argmin_{z_1, ..., z_n} ~ \sum_{k=1}^K \sum_{x_i, x_j : z_i = z_j = k}  d^2(x_i, x_j).
$$
Here, $d^2(x_i, x_j)$ is the squared distance between objects $x_i$ and $x_j$. 
If they are vectors in Euclidean space, then we can use $d^2(x_i, x_j) = \|x_i - x_j\|_2^2$, 
or if they are vertices on a graph, then $d^2(x_i, x_j)$ represents some notion of graph distance. 
While this formulation of $K$-means clustering takes care of the lack of centroids, the implementation still requires some thought since Lloyd's algorithm includes the computation of centroids. 
An alternative algorithm for $K$-means clustering in this setting is MacQueen's exchange algorithm \citep{MacQueen1967}, which involves cycling through each vertex and then, for each vertex, cycling through the labels and choosing the label that minimizes the objective function.

Yet another approach to adapting $K$-means clustering to graph data is to embed the graph to Euclidean space and then apply Lloyd's algorithm on the embedding. 
In this context, a natural choice of embedding is one in which pairwise Euclidean distances between embedding vectors are (approximately) equal to pairwise graph distances between corresponding vertices. 
For example, the combinatorial Laplacian eigenmap is an embedding in which Euclidean distances between pairs of vectors in the embedding are equal to expected commute times between corresponding pairs of vertices in the graph \citep{vonLuxburg2007}. 
This embedding is constructed by computing the combinatorial Laplacian matrix $L = D - A$ in which $D$ is a diagonal matrix in which each $D_{ii}$ is the degree of vertex $i$. 
Then the components of the embedding consist of $\big[ v_{n-1} / \sqrt{\lambda_{n - 1}} \mid \cdots \mid v_1 / \sqrt{\lambda_1} \big]^\top$ in which $\lambda_i$ is the $i^{th}$ largest eigenvalue of $L$ and $v_i$ is its corresponding eigenvector ($\lambda_n = 0$ and is omitted). 
Then Lloyd's algorithm can be applied to this embedding. 
More often in practice, the $d$ smallest eigenvalues are used to construct a lower dimensional embedding. 
\end{example}

@@

Given a sample or dataset, a typical analysis task is statistical inference, or the estimation of various parameters under the assumption that the data come from a random distribution or process. 
These estimated parameters are often then used for making predictions or deriving insights about the population. 
For example, when fitting a Gaussian mixture model, the data are first assumed to come from a mixture of Gaussians. 
The model fitting process then involves estimating the means and standard deviations of each Gaussian component, along with the mixture weights. 
The resulting model provides insight into where each mixture component is located, how disperse each component is, and how the data are distributed between the components.  
In order to perform a similar type of analysis for graphs, we must first define probability distributions from which such data can be sampled.

@@

Statistical inference is not possible on a general Bernoulli random graph with arbitrary edge probabilities since the number of parameters (individual edge probabilities) is equal to the number of observations (presence or absence of an edge). 
Additional structure must be introduced. 
One such structured Bernoulli random graph model is the $\ER$ graph, first proposed by \cite{Gilbert:1959}, which is defined as follows:

@@

In this work, we explore three types of block models, the stochastic block model, the degree corrected block model, and the popularity adjusted block model, as well as a family of latent space models called generalized random dot product graphs.

The contributions of this work are as follows. 
In section \@ref(sec:pabm-grdpg), we show that, similar to how the stochastic block model and degree corrected block model are generalized random dot product graphs with specific latent structures, the popularity adjusted block model is also a generalized random dot product graph with a specific latent structure. 
More specifically, we show that the popularity adjusted block model is a generalized random dot product graph in which the latent vectors are organized by community and each community in the latent space is defined by a $K$-dimensional subspace, where $K$ is the number of communities in the model. 
We then exploit this rigid and linear structure for community detection by applying sparse subspace clustering on the adjacency spectral embedding of the graph to infer the community labels. 
We also propose a novel algorithm called orthogonal spectral clustering, which constructs an embedding for the popularity adjusted block model which forces the communities to lie on *orthogonal* subspaces. 
Finally, we propose a spectral decomposition based parameter estimation algorithm for the popularity adjusted block model. 
All three algorithms have consistency properties that result in zero error with increasing graph size.

In section \@ref(sec:mbm), we extend the latent structures of the stochastic block model, degree corrected block model, and popularity adjusted block model to nonlinear latent structures. 
Instead of points, line segments, and subspaces, we propose the manifold block model which is a class of generalized random dot product graph in which the latent vectors lie on one of $K$ manifolds. 
In the case of nonintersecting manifolds, we show that a clustering algorithm based on a nearest algorithms graph approach on an embedding is consistent, resulting in zero clustering error for large graphs, as long as the latent vectors are sampled from the manifolds densely 
In the case of intersecting manifolds, we propose the $K$-curves algorithm, which is a type of batch coordinate ascent algorithm which alternates between fitting manifolds to the embedding and reassigning clusters based on proximity to the manifolds.

We conclude with section \@ref(sec:comparison) in which we compare the four block models.

@@

The stochastic block model (SBM) \citep{doi:10.1080/0022250X.1971.9989788} is the simplest of the three block models and assumes that each pair of communities $(k, \ell)$ has a constant edge probability $\theta_{k \ell}$. 
We now give the formal definition of the SBM within the context of Bernoulli random graphs:

@@

\begin{example}[Stochastic block model with $K = 2$ communities]
\label{ex:assort-sbm}
We construct an SBM with two communities in which $\theta_{11} = 1/2$, $\theta_{12} = 1/8$, and $\theta_{22} = 1/4$. 
Note that in this example, the within-community edge probabilities are greater than the between-community edge probability. 
A realization of this graph with $n_1 = n_2 = 32$ (where $n_k = |\{v_i : z_i = k\}|$) is illustrated in figure \ref{fig:assort-sbm}.

@@

\begin{example}[Dating network as a disassortative stochastic block model]
\label{ex:sbm-dating}
Consider an undirected graph in which each vertex $v_i$ represents users of an online dating service, each label $z_i$ represents the user's gender, and each edge represents a successful match between pairs of users. 
For simplicity, we will restrict the labels to female and male (denoted as $1$ and $2$). 
If we model this as an SBM, then $\theta_{11}$ is the probability of a match between two female users, $\theta_{22}$ is the probability of a match between two male users, and $\theta_{12} = \theta_{21}$ is the probability of a match between a female user and a male user. 
Based on trends in the United States \citep{gallup-lgbt}, 
a plausible set of edge probabilities is $\theta_{11} = \theta_{22} = 0.02$ and $\theta_{12} = 0.2$, which are used in the sample visualized in figure \ref{fig:dating-sbm}, along with $\alpha_1 = \alpha_2 = 1/2$ and $n = 64$.

@@

\begin{equation}
\label{eq:sbm-eigenvalues}
\lambda = \frac{n}{2} \bigg( 
\theta_{11} + \theta_{22} \pm 
\sqrt{(\theta_{11} - \theta_{22})^2 + 4 \theta_{12}^2} 
\bigg).
\end{equation}
The first eigenvalue is always positive. 
The second eigenvalue is positive if $\theta_{11} \theta_{22} > \theta_{12}^2$. 
Thus, $P$ describes an assortative SBM if $\theta_{11} \theta_{22} > \theta_{12}^2$ or 
a disassortative SBM if $\theta_{11} \theta_{22} < \theta_{12}^2$. 
This corresponds to the high-level understanding of assortativity and disassortativity based on whether within or between community edge probabilities are greater. 
\end{example}

@@

The SBM is a type of mixture model, and like most classical mixture models and clustering problems, likelihood maximization is NP-hard. 
But because the SBM is a mixture model, a candidate algorithm for finding a (local) maximum of the likelihood is the widely studied expectation maximization (EM) algorithm \citep{10.2307/2984875}.
However, due to the $z_{ik} z_{j \ell}$ interaction terms, the expectation step cannot be solved in closed form \citep{kolaczyk2014statistical}. 
While the labels are drawn independently, they are not necessarily conditionally independent. 
Nevertheless, making the conditional independence relaxation allows us to write the expectation and maximization steps in closed form. 
(The same type of approximation was used in the Bayesian context by \citet{10.2307/43974715}.) 
As an example, we write the EM algorithm for the likelihood function in equation \eqref{eq:sbm-loglik} in algorithm \ref{algo:sbm-em}.

@@

In many real networks, the degree of each vertex often does not grow proportionally with the size of the network. 
To account for this, a sparsity factor $\rho_n \in (0, 1]$ is introduced in the edge probabilities, i.e., $P_{ij} \leftarrow \rho_n P_{ij}$, for some sequence $\{\rho_n\}$ such that $\lim\limits_{n \to \infty} \rho_n = 0$.  
For example, a sparse SBM has edge probabilities $P_{ij} = \rho_n \theta_{z_i, z_j}$, for which we use the notation $A \sim \SBM(z, \{\theta_{k \ell}\}_K; \rho_n)$ or $A \sim \SBM(\alpha, \{\theta_{k \ell}\}_K; \rho_n)$, depending on whether we treat the labels as random or fixed. 
Then the expected degree grows as $O(n \rho_n)$ instead of linearly as $O(n)$. 
For the sake of unifying the sparse and dense regimes, we also allow for the special case $\rho_n = 1$ and include the sparsity factor throughout, unless otherwise stated. 
Finally, we also note that while $\rho_n$ limits the rate of growth of the expected degree, our theoretical results still require $n \rho_n$ to diverge to infinity, albeit at a slower rate than $O(n)$. 
In general, for consistency in estimation and inference, most results on Bernoulli random graphs require $n \rho_n = \omega \big( (\log n)^c \big)$ for some $c > 1$. 
(See \cite{JMLR:v18:16-480}, \cite{https://doi.org/10.48550/arxiv.2106.09840}, and \cite{rubindelanchy2017statistical} for further discussion.)

@@

The SBM assumes homogeneity within each community. 
In particular, in the SBM, each vertex's expected degree and affinity toward each community depends only on its community label. 
The degree corrected block model (DCBM) and popularity adjusted block model (PABM) are generalizations of the SBM that address these limitations. 
In particular, the DCBM allows each vertex to have an expected degree independent of its label, and the PABM allows each vertex to have edge probabilities to each community independent of its own label. 
We also show that the DCBM is a special case of the PABM and that the SBM, DCBM, and PABM are nested models.

\begin{definition}[Degree corrected block model]
\label{def:dcbm}
Let $G = (V, E)$ be a Bernoulli random graph with $n$ vertices, described by random adjacency matrix $A$. 
Let $K \geq 1$ be an integer that describes the number of communities and $\theta_{k \ell} \in [0, 1]$ for each $k, \ell \in \{1, ..., K\}$ (if $G$ is undirected, then $\theta_{k \ell} = \theta_{\ell k}$), as in the SBM. 
Let $z_1, ..., z_n \in \{1, ..., K\}$ be community labels associated with each vertex. 
In addition, each vertex $v_i$ has a degree correction parameter $\omega_i \in [0, 1]$. 
If the edge probability matrix $P$ for this graph is such that $P_{ij} = \rho_n \theta_{z_i, z_j} \omega_i \omega_j$ and $A \sim \BG(P)$, then $G$ is a degree corrected block model. 
We use the notation $A \sim \DCBM(z, \{\theta_{k \ell}\}_K, \omega; \rho_n)$ to denote a random adjacency matrix drawn from the DCBM with labels $z$, base edge probabilities $\{\theta_{k \ell}\}_K$ (of which there are $K (K+1) / 2$), and degree correction factors $\omega \in \mathbb{R}^n$. 
Under the sampling scheme in which the labels are drawn from a multivariate distribution with class probabilities $\alpha$, we use the notation $A \sim \DCBM(\alpha, \{\theta_{k \ell}\}_K, \omega; \rho_n)$.
\end{definition}

\begin{example}[Dating network as a disassortative DCBM]
\label{ex:dcbm-dating}
In example \ref{ex:sbm-dating}, we modeled an online dating network as an SBM in which the vertices correspond to individuals, the communities correspond to genders, and edges correspond to whether the online dating service successfully matched pairs of individuals. 
In this model, there is an implicit assumption that for each pair of genders, there is a constant probability of a match, e.g., each male member has a constant probability of matching with another male member as well as a (different) constant probability of matching with a female member. 
This does not take into account that some individuals are more likely to form connections than others, based on each individual's activity on the online service or popularity with other users. 
Modeling this as a DCBM allows us to take this into account by adjusting each user's expected degree by a factor of $\omega_i$. 
Figure \ref{fig:dating-dcbm} is a visualization of the network from example \ref{ex:sbm-dating} modified as a DCBM with parameters $\theta_{11} = \theta_{22} = 0.03$, $\theta_{12} = 0.3$, each $\omega_i$ drawn uniformly between $1/3$ and $1$, and $\rho_n = 1$. 
The figure shows that some individuals (vertices) are a lot more popular than others in the form of having more matches with other individuals (edges).

@@

The DCBM allows for each vertex to have its own expected degree while still maintaining community structure. 
However, it still assumes that each vertex within a community has the same relative propensity to connect to each community. 
Referring again to the dating network example (examples \ref{ex:sbm-dating} and \ref{ex:dcbm-dating}), the model assume that each individual is ten times more likely to connect with a member of the opposite sex than with a member of the same sex, without regard to each individual's preferences. 
The popularity adjusted block model \citep{307cbeb9b1be48299388437423d94bf1} allows each vertex to have its own affinity toward each community. 
In the dating network example, the PABM allows us to model each individual's sexual orientation instead of using population level averages.

@@

In the Bernoulli random graph, the edge probability matrix $P$ is of size $n \times n$, but when $K \ll n$, the structure of $P$ is much simpler than its dimensions imply. 
In particular, $P$ is rank $K$ in the SBM and DCBM and rank $K^2$ in the PABM. 
This property lends these block models to be analyzed as another family of Bernoulli random graphs with low rank called the random dot product graph (RDPG) and generalized random dot product graph (GRDPG). 
In these models, $P$ is decomposed into an outer product of two low-rank data matrices.

@@

We use the notation $A \sim \GRDPG_{p,q}(X; \rho_n)$ to denote a random adjacency matrix $A$ drawn from latent positions $X$, sparsity parameter $\rho_n$, and signature $(p, q)$. If the $n$ vectors in $X$ are drawn from probability distribution $F$ on support $\mathcal{X}$, we use the notation $A \sim \GRDPG_{p,q}(F; \rho_n)$.

>>>>

Graph and network data have become increasingly widespread in various fields including sociology, neuroscience, biostatistics, and computer science. 
This has resulted in challenges for researchers who rely on traditional statistical and machine learning methods that are incompatible with graph data and instead assume that the data exist as feature vectors. 
To illustrate this, consider the typical approach to building a statistical or machine learning model. 
Data are often represented as an $n \times d$ matrix $X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top$ in which each row $x_i \in \mathbb{R}^d$ is an observation of $d$ features and each column is a set of $n$ feature measurements. 
An analysis task for these data might be to come up with a classification model $\phi : \mathbb{R}^d \to \{1, 2, ..., K\}$ that uses the numerical values of each feature of a vector $x_i$ to calculate a predicted label $z_i \in \{1, 2, ..., K\}$. 
For instance, $\phi(x)$ might first compute the distances from $x$ to $K$ points in $\mathbb{R}^d$ and then assign $x$ to the label of the nearest point. 
Examples of this include linear discriminant analysis (in the case of supervised learning) and Lloyd's algorithm \citep{1056489} or Gaussian mixture models \citep{doi:10.1198/016214502760047131} (in the case of unsupervised learning). 
However, it is not obvious how this method would translate to data that are represented as graphs, in which observations consist of relationships among a set of objects rather than numerical attributes associated with each object: 
Instead of feature vector $x_i = \Big[ x_{i1} \mid \cdots \mid x_{id} \Big]^\top \in \mathbb{R}^d$, we observe $a_i = \Big[ a_{i1} \mid \cdots \mid a_{in} \Big]^\top \in \mathbb{R}^n$ in which each $a_{ij}$ is object $i$'s relationship to object $j$. 
For such data, it is often necessary to alter existing algorithms, transform the graph data into Euclidean data, often called *graph embedding* or *spectral clustering* \citep{vonLuxburg2007}, and apply algorithms for Euclidean data on the embedding, or come up with new algorithms altogether.

\begin{example}[Graph clustering]
$K$-means clustering \citep{MacQueen1967} is a nonparametric clustering method that assigns labels $z_1, ..., z_n \in \{1, ..., K\}$ to feature vectors $x_1, ..., x_n \in \mathbb{R}^d$ by minimizing the objective function
\begin{equation}
\label{eq:kmeans}
\argmin_{z_1, ..., z_n} ~ W(z_1, ..., z_n; x_1, ..., x_n) = 
\argmin_{z_1, ..., z_n} ~ \sum_{k = 1}^K \sum_{x_i : z_i = k} \|x_i - \bar{x}_k\|_2^2.
\end{equation}
$\bar{x}_1, ..., \bar{x}_K \in \mathbb{R}^p$ are the centroids of each cluster, which act as nuisance parameters. 
A popular implementation of $K$-means clustering is Lloyd's algorithm \citep{1056489}, which is a type of coordinate descent algorithm in which at each iteration the cluster labels are updated according to the label of its nearest centroid and centroids are updated as the sample means of the points in each cluster.

If we have a graph instead of vectors in $\mathbb{R}^d$, it is not obvious how to translate $K$-means clustering to these data. 
In particular, there is no inherent notion of centroid or sample mean for graph data. 
However, there are inherent notions of distances among vertices, such as shortest path distance, expected commute time, or resistance distance. 
One way to adapt $K$-means to graph data is to use an equivalent objective function as equation \eqref{eq:kmeans} that doesn't use centroids (which are just nuisance parameters) and is stated only in terms of distances:  
$$
\argmin_{z_1, ..., z_n} ~ \tilde{W}(z_1, ..., z_n; x_1, ..., x_n) =
\argmin_{z_1, ..., z_n} ~ \sum_{k=1}^K \sum_{x_i, x_j : z_i = z_j = k}  d^2(x_i, x_j).
$$
Here, $d^2(x_i, x_j)$ is the squared distance between objects $x_i$ and $x_j$. 
If they are vectors in Euclidean space, then we can use $d^2(x_i, x_j) = \|x_i - x_j\|_2^2$, 
or if they are vertices on a graph, then $d^2(x_i, x_j)$ represents some notion of graph distance. 
While this formulation of $K$-means clustering takes care of the lack of centroids, the implementation still requires some thought since Lloyd's algorithm includes the computation of centroids. 
An alternative algorithm for $K$-means clustering in this setting is MacQueen's exchange algorithm \citep{MacQueen1967}, which involves cycling through each vertex and then, for each vertex, cycling through the labels and choosing the label that minimizes the objective function.

Yet another approach to adapting $K$-means clustering to graph data is to embed the graph to Euclidean space and then apply Lloyd's algorithm on the embedding. 
In this context, a natural choice of embedding is one in which pairwise Euclidean distances between embedding vectors are (approximately) equal to pairwise graph distances between corresponding vertices. 
For example, the combinatorial Laplacian eigenmap is an embedding in which Euclidean distances between pairs of vectors in the embedding are equal to expected commute times between corresponding pairs of vertices in the graph \citep{vonLuxburg2007}. 
This embedding is constructed by computing the combinatorial Laplacian matrix $L = D - A$ in which $D$ is a diagonal matrix in which each $D_{ii}$ is the degree of vertex $i$. 
Then the components of the embedding consist of $\big[ v_{n-1} / \sqrt{\lambda_{n - 1}} \mid \cdots \mid v_1 / \sqrt{\lambda_1} \big]^\top$ in which $\lambda_i$ is the $i^{th}$ largest eigenvalue of $L$ and $v_i$ is its corresponding eigenvector ($\lambda_n = 0$ and is omitted). 
Then Lloyd's algorithm can be applied to this embedding. 
More often in practice, the $d \ll n$ smallest eigenvalues are used to construct a lower dimensional embedding. 
\end{example}

@@

One approach to analyzing graphs as network data is statistical inference, or the estimation of various parameters under the assumption that the data come from a random distribution or process. 
These estimated parameters are often then used for making predictions or deriving insights about the population. 
For example, when fitting a Gaussian mixture model, the data are first assumed to come from a mixture of Gaussians. 
The model fitting process then involves estimating the means and standard deviations of each Gaussian component, along with the mixture weights. 
The resulting model provides insight into where each mixture component is located, how disperse each component is, and how the data are distributed between the components. 
In order to perform a similar type of analysis for graphs, we must first define probability distributions from which such data can be sampled.

@@

Statistical inference is not possible on a general Bernoulli random graph with arbitrary edge probabilities since the number of parameters (individual edge probabilities) is equal to the number of observations (presence or absence of an edge). 
Additional structure must be introduced. 
One such structured Bernoulli random graph model is the $\ER$ graph, first proposed by \cite{Gilbert:1959}, which restricts every edge probability to a constant value.

@@

In this work, we explore three types of block models, the stochastic block model, the degree corrected block model, and the popularity adjusted block model, as well as a family of latent space models called generalized random dot product graphs. 
In particular, this work explores the ways in which block models and latent space models are related and uses that connection to develop methods for analyzing and fitting models to graph data. 
The contributions of this work are as follows. 
In section \@ref(sec:pabm-grdpg), we show that, similar to how the stochastic block model and degree corrected block model are generalized random dot product graphs with specific latent structures, the popularity adjusted block model is also a generalized random dot product graph with a specific latent structure. 
More specifically, we show that the popularity adjusted block model is a generalized random dot product graph in which the latent vectors are organized by community and each community in the latent space is defined by a $K$-dimensional subspace, where $K$ is the number of communities in the model. 
We then exploit this rigid and linear structure for community detection by applying sparse subspace clustering on the adjacency spectral embedding of the graph to infer the community labels. 
We also propose a novel algorithm called orthogonal spectral clustering, which constructs an embedding for the popularity adjusted block model which forces the communities to lie on *orthogonal* subspaces. 
Finally, we propose a spectral decomposition based parameter estimation algorithm for the popularity adjusted block model. 
All three algorithms have consistency properties that result in zero error with increasing graph size.

In section \@ref(sec:mbm), we extend the latent structures of the stochastic block model, degree corrected block model, and popularity adjusted block model to nonlinear latent structures. 
Instead of points, line segments, and subspaces, we propose the manifold block model which is a class of generalized random dot product graph in which the latent vectors lie on one of $K$ manifolds. 
In the case of nonintersecting manifolds, we show that a clustering algorithm based on a nearest algorithms graph approach on an embedding is consistent, resulting in zero clustering error for large graphs, as long as the latent vectors are sampled from the manifolds densely 
In the case of intersecting manifolds, we propose the $K$-curves algorithm, which is a type of batch coordinate ascent algorithm which alternates between fitting manifolds to the embedding and reassigning clusters based on proximity to the manifolds. 
We conclude with section \@ref(sec:comparison) in which we compare the four block models and summarize the ways in which one might determine which block model is most appropriate for one's data or setting.

@@

The stochastic block model (SBM) \citep{doi:10.1080/0022250X.1971.9989788} is the simplest of the three block models and assumes that each pair of communities $(k, \ell)$ has a constant edge probability $\theta_{k \ell}$. 
The formal definition of the SBM within the context of Bernoulli random graphs is as follows:

@@

\begin{example}[Stochastic block model with $K = 2$ communities]
\label{ex:assort-sbm}
In this SBM, there are two communities with parameters $\theta_{11} = 1/2$, $\theta_{12} = 1/8$, and $\theta_{22} = 1/4$. 
Note that in this example, the within-community edge probabilities are greater than the between-community edge probability. 
A realization of this graph with $n_1 = n_2 = 32$ (where $n_k = |\{v_i : z_i = k\}|$) is illustrated in figure \ref{fig:assort-sbm}.

@@

\begin{example}[Dating network as a disassortative stochastic block model]
\label{ex:sbm-dating}
Consider an undirected graph in which each vertex $v_i$ represents a user of an online dating service, each label $z_i$ represents the user's gender, and each edge represents a successful match between pairs of users. 
For simplicity, we will restrict the labels to female and male (denoted as $1$ and $2$, respectively). 
If we model this as an SBM, then $\theta_{11}$ is the probability of a match between two female users, $\theta_{22}$ is the probability of a match between two male users, and $\theta_{12} = \theta_{21}$ is the probability of a match between a female user and a male user. 
Based on trends in the United States \citep{gallup-lgbt}, 
a plausible set of edge probabilities is $\theta_{11} = \theta_{22} = 0.02$ and $\theta_{12} = 0.2$, which are used in the sample visualized in figure \ref{fig:dating-sbm}, along with $\alpha_1 = \alpha_2 = 1/2$ and $n = 64$.

@@

\begin{equation}
\label{eq:sbm-eigenvalues}
\lambda = \frac{n}{2} \bigg( 
\theta_{11} + \theta_{22} \pm 
\sqrt{(\theta_{11} - \theta_{22})^2 + 4 \theta_{12}^2} 
\bigg).
\end{equation}
The first eigenvalue is always positive. 
The second eigenvalue is positive if $\theta_{11} \theta_{22} > \theta_{12}^2$. 
Thus, $P$ describes an assortative SBM if $\theta_{11} \theta_{22} > \theta_{12}^2$ or 
a disassortative SBM if $\theta_{11} \theta_{22} < \theta_{12}^2$. 
This corresponds to the high-level understanding of assortativity and disassortativity based on whether within community edge probabilities are greater than the between community edge probabilities. 
\end{example}

@@

The SBM is a type of mixture model, and like most classical mixture models and clustering problems, likelihood maximization is NP-hard. 
But because the SBM is a mixture model, a candidate algorithm for finding a (local) maximum of the likelihood is the widely studied expectation maximization (EM) algorithm \citep{10.2307/2984875}.
However, due to the $z_{ik} z_{j \ell}$ interaction terms, the expectation step cannot be solved in closed form \citep{kolaczyk2014statistical}. 
While the labels are drawn independently, they are not necessarily conditionally independent. 
Nevertheless, making the conditional independence relaxation allows us to write the expectation and maximization steps in closed form. 
(The same type of approximation was used in the Bayesian context by \citet{10.2307/43974715}.) 
The EM algorithm for maximizing the likelihood function in equation \eqref{eq:sbm-loglik} is detailed Kin algorithm \ref{algo:sbm-em}.

@@

In many real networks, the degree of each vertex often does not grow proportionally with the size of the network. 
To account for this, a sparsity factor $\rho_n \in (0, 1]$ is introduced in the edge probabilities, i.e., $P_{ij} \leftarrow \rho_n P_{ij}$, for some sequence $\{\rho_n\}$. 
Oftentimes the additional constraint of $\lim\limits_{n \to \infty} \rho_n = 0$ is included. 
For example, a sparse SBM has edge probabilities $P_{ij} = \rho_n \theta_{z_i, z_j}$, for which we use the notation $A \sim \SBM(z, \{\theta_{k \ell}\}_K; \rho_n)$ or $A \sim \SBM(\alpha, \{\theta_{k \ell}\}_K; \rho_n)$, depending on whether we treat the labels as random or fixed. 
Then the expected degree grows as $O(n \rho_n)$ instead of linearly as $O(n)$. 
For the sake of unifying the sparse and dense regimes, we also allow for the special case $\rho_n = 1$ and include the sparsity factor throughout, unless otherwise stated. 
Finally, we also note that while $\rho_n$ limits the rate of growth of the expected degree, our theoretical results still require $n \rho_n$ to diverge to infinity, albeit at a slower rate than $O(n)$. 
For example, if $\rho_n \propto 1 / \sqrt{n}$, then the expected degree for each vertex of the SBM is $O(\sqrt{n})$. 
In general, for consistency in estimation and inference, most results on Bernoulli random graphs require $n \rho_n = \omega \big( (\log n)^c \big)$ for some $c > 1$. 
(See \cite{JMLR:v18:16-480}, \cite{https://doi.org/10.48550/arxiv.2106.09840}, and \cite{rubindelanchy2017statistical} for further discussion.)

@@

The SBM assumes homogeneity within each community. 
More specifically, in the SBM, each vertex's expected degree and affinity toward each community depends only on its community label. 
The degree corrected block model (DCBM) and popularity adjusted block model (PABM) are generalizations of the SBM that address these limitations. 
In particular, the DCBM allows each vertex to have an expected degree independent of its label, and the PABM allows each vertex to have edge probabilities to each community independent of its own label. 
We also show that the DCBM is a special case of the PABM and that the SBM, DCBM, and PABM are nested models.

\begin{definition}[Degree corrected block model]
\label{def:dcbm}
Let $G = (V, E)$ be a Bernoulli random graph with $n$ vertices, described by random adjacency matrix $A$. 
Let $K \geq 1$ be an integer that describes the number of communities and $\theta_{k \ell} \in [0, 1]$ for each $k, \ell \in \{1, ..., K\}$ (if $G$ is undirected, then $\theta_{k \ell} = \theta_{\ell k}$) be the baseline community-wise edge probabilities, as in the SBM. 
Let $z_1, ..., z_n \in \{1, ..., K\}$ be community labels associated with each vertex. 
In addition, each vertex $v_i$ has a degree correction parameter $\omega_i \in [0, 1]$. 
If the edge probability matrix $P$ for this graph is such that $P_{ij} = \rho_n \theta_{z_i, z_j} \omega_i \omega_j$ and $A \sim \BG(P)$, then $G$ is a degree corrected block model. 
We use the notation $A \sim \DCBM(z, \{\theta_{k \ell}\}_K, \omega; \rho_n)$ to denote a random adjacency matrix drawn from the DCBM with labels $z$, base edge probabilities $\{\theta_{k \ell}\}_K$ (of which there are $K (K+1) / 2$), and degree correction factors $\omega \in \mathbb{R}^n$. 
Under the sampling scheme in which the labels are drawn from a multivariate distribution with class probabilities $\alpha$, we use the notation $A \sim \DCBM(\alpha, \{\theta_{k \ell}\}_K, \omega; \rho_n)$.
\end{definition}

\begin{example}[Dating network as a disassortative DCBM]
\label{ex:dcbm-dating}
In example \ref{ex:sbm-dating}, we modeled an online dating network as an SBM in which the vertices correspond to individuals, the communities correspond to genders, and edges correspond to whether the online dating service successfully matched pairs of individuals. 
In this model, there is an implicit assumption that for each pair of genders, there is a constant probability of a match, e.g., each male member has a constant probability of matching with any other male member as well as a (different) constant probability of matching with any female member. 
This does not take into account that some individuals are more likely to form connections than others, based on each individual's activity on the online service or popularity with other users. 
Modeling this as a DCBM allows us to take this into account by adjusting each user's expected degree by a factor of $\omega_i$. 
Figure \ref{fig:dating-dcbm} is a visualization of the network from example \ref{ex:sbm-dating} modified as a DCBM with parameters $\theta_{11} = \theta_{22} = 0.03$, $\theta_{12} = 0.3$, each $\omega_i$ drawn uniformly between $1/3$ and $1$, and $\rho_n = 1$. 
The figure shows that some individuals (vertices) are a lot more popular than others in the form of having more matches with other individuals (edges).

@@

The DCBM allows for each vertex to have its own expected degree while still maintaining community structure. 
However, it still assumes that each vertex within a community has the same relative propensity to connect to each community. 
Referring again to the dating network examples  \ref{ex:sbm-dating} and \ref{ex:dcbm-dating}, the model assume that each individual is ten times more likely to connect with a member of the opposite sex than with a member of the same sex, without regard to each individual's sexual orientation. 
The popularity adjusted block model \citep{307cbeb9b1be48299388437423d94bf1} allows each vertex to have its own affinity toward each community. 
In the dating network example, the PABM allows us to model each individual's sexual orientation instead of using population level averages.

@@

In block models, the edge probability matrix $P$ is of size $n \times n$, but when $K \ll n$, the structure of $P$ is much simpler than its dimensions imply. 
In particular, $P$ is rank $K$ in the SBM and DCBM and rank $K^2$ in the PABM. 
This property lends these block models to be analyzed as another family of Bernoulli random graphs with low rank called the random dot product graph (RDPG) and generalized random dot product graph (GRDPG). 
In these models, $P$ is decomposed into an outer product of two low-rank data matrices.

@@

We use the notation $A \sim \GRDPG_{p,q}(X; \rho_n)$ to denote a random adjacency matrix $A$ drawn from latent positions $X$, sparsity parameter $\rho_n$, and signature $(p, q)$. If the $n$ latent vectors in $X$ are drawn from probability distribution $F$ on support $\mathcal{X}$, we use the notation $A \sim \GRDPG_{p,q}(F; \rho_n)$.
